---
title: "ANOVA Exercises"
subtitle: ""  
author: 
  - "AJ Smit"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css:
      - default
      - "default-fonts.css"
      - ki
      - "uwc_logo.css"
    yolo: false
    self_contained: TRUE  
    lib_dir: libs
    nature:
      highlightStyle: idea
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      ratio: 16:9
---

```{r setup, include=FALSE, echo=FALSE} 
library(tidyverse)
library(ggpubr)
knitr::opts_chunk$set(echo = TRUE,
                      comment = "R>",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      fig.retina = 3,
                      fig.width = 6,
                      fig.height = 3.6,
                      fig.align = "center")
```

# Question 1

* Here is bunch of data for pigs raised on different diets. The experiment is similar to the chicken one. Does feed type have an effect on the mass of pigs at the end of the experiment?

```{r, eval=FALSE}
# enter the mass at the end of the experiment
feed_1 <- c(60.8, 57.0, 65.0, 58.6, 61.7)
feed_2 <- c(68.7, 67.7, 74.0, 66.3, 69.8)
feed_3 <- c(102.6, 102.1, 100.2, 96.5)
feed_4 <- c(87.9, 84.2, 83.1, 85.7, 90.3)

# make a dataframe
bacon <- as.tibble(data.frame(
  feed = c(
  rep("Feed 1", length(feed_1)),
  rep("Feed 2", length(feed_2)),
  rep("Feed 3", length(feed_3)),
  rep("Feed 4", length(feed_4))
  ),
  mass = c(feed_1, feed_2, feed_3, feed_4)
  ))
```

---

.left-column[
# Question 2
]

.right-column[
* Construct suitable null and alternative hypotheses for the built-in `ToothGrowth` data, and test your hypotheses using an ANOVA.
]

---

.left-column[
# Question 3
]

.right-column[
* Find or generate your own data that lend themselves to being analysed by a two-way ANOVA. Generate suitable hypotheses about your data, and analyse it. Supplement your analysis by providing a suitable descriptive statistical summary and graph(s) of your data.
]
